A Comparison of Parallel Graph Processing Implementations

### 2017-05-16 papers/publication/paper-pollard.tex
Added todo for redoing graphalytics table.
First submission to Cluster. Rewrote, reorganized.
Still waiting on scale 23 data.

An overarching issue among these systems is the lack of comprehensive comparisons. One possible reason is the considerable effort involved in getting each system to run: satisfying dependencies and ensuring data are correctly formatted can be time consuming tasks. Moreover, some systems are developed with large, multi-node clusters in mind while others only work with shared memory, single-node computers. An example of the former is GraphX~\cite{Xin:2013:GraphX}, which incurs some overhead while gaining fault tolerance and an example of the latter is Ligra~\cite{Shun:2013:Ligra}, a framework requiring a shared-memory architecture.
% ``The project was motivated by the fact that the largest publicly available real-world graphs all fit in shared memory''

@@

To address these issues, we introduce \mbox{\emph{easy-parallel-graph-\textasteriskcentered}},\footnote{Available at \url{https://github.com/HPCL/easy-parallel-graph}}---a framework that simplifies the installation, comparison, and performance analysis of the three most widely implemented graph algorithm building blocks: breadth first search (BFS), single source shortest paths (SSSP), and PageRank (PR). With this framework we select a small number of graph processing libraries over which we homogenize all aspects of execution to compare performance fairly. Another contribution of this paper is a demonstration of the use of this framework to analyze the performance, energy consumption, and scalability of specific algorithm implementation. This enables algorithm designers to select high performance reference implementations with which they can compare their improvements. To be clear, we are not proposing a new benchmark suite or providing any new reference implementations. Instead, we introduce a method for detailed and fair comparison of existing graph software packages without requiring the user to be familiar with them.

\section{Related Work}\label{sec:relatedwork}
Most performance analysis of graph processing systems come from the empirical results of each new library designer's publications. For example, GraphMat~\cite{Sundaram:2015:GraphMat}, PowerGraph~\cite{Gonzalez:2012:Powergraph}, and GraphBIG~\cite{Nai:2015:Graphbig} present a comparison of performance between their system and a selection of other software packages and in general any new approach will compare itself to existing implementations. These performance results are inherently biased because the authors typically know their own package better than the competition and as such are aware of more optimizations.

Nai et al.~\cite{Nai:2016:architectural} provide a detailed performance analysis using GraphBIG as their reference implementation. Their approach also compares GraphBIG to the GraphLab and Pregel execution models (Gather Apply Scatter). Their analysis considers architectural performance measurements such as cache miss rates to measure bottlenecks and computational efficiency for a variety of datasets. While incredibly thorough, these analyses focus on covering a wide range of datasets and appear difficult to apply to a new approach. For example, GraphMat reduces computation to sparse matrix operations which may not suffer from the same memory bottlenecks indicated in the results of Nai et al.

Research by Song et al.~\cite{Song:2016:LoadBalancing} and LeBeane et al.~\cite{LeBeane:2015:GraphHetero} considers the important problem of partitioning on heterogeneous architectures and provide profiling to reduce runtime, improve load balance, and reduce energy consumption.  Our framework instead focuses on shared memory packages.

Beyond the aforementioned performance results, a number of reports focus on performance analysis not tied to a particular software package. Satish et al.~\mbox{\cite{Satish:2014:NavigatingGraph}} analyze the performance of several systems on datasets on the order of $30$ billion edges and~\cite{Lu:2014:ExperimentalEval} uses six real-world datasets and focuses on the vertex-centric programming model. These implementations are hand tuned and provides recommendations for future improvements.

@@

\begin{table}
	\caption{Graphalytics: tabulated sample run times (seconds) with 32 threads. Just one run per experiment is performed. Below the table is an excerpt from the GraphMat log file.}
		\label{tab:graphalytics-realworld}
	\vspace{-10pt}
	\begin{center}
	% cit-Patents: (3 774 768 vertices, 16 518 948 edges)
	% dota-league (61 170 vertices, 50 870 313 edges)
	\begin{tabular}{l|c|c|c|c|c|c}
	%% BN: Commented out because it's redundant with the caption
	%\multicolumn{5}{l}{\large{\emph{Graphalytics}}}  \\
	GraphBIG & BFS  & CDLP  &   LCC   &  PR &  SSSP & WCC \\ \hline
	cit-Patents & 0.8 & 11.8 & 15.5  &  4.5 & N/A  & 1.3 \\
	dota-league & 1.1 & 3.9 & 1073.7 &  2.6 & \textbf{3.0} & 1.0 \\
	\multicolumn{1}{l}{}\\
	PowerGraph & BFS  & CDLP  &   LCC   &  PR &  SSSP & WCC \\ \hline
	cit-Patents & 13.8 & 30.1 & 23.9 & 18.8 & N/A & 22.1 \\
	dota-league & 25.6 & 31.2 & 458.1 & 26.7 & 28.9 & 22.9 \\
	% graph500-22 43.0 s 55.6 s 299.8 s 46.4 s N/A    40.5 s
	\multicolumn{1}{l}{}\\
	GraphMat & BFS  & CDLP  &   LCC   &  PR &  SSSP & WCC \\ \hline
	cit-Patents & 7.8 & 20.6 & 11.2 &  8.4 & N/A &  7.0 \\
	dota-league & 2.2 & 19.7 & 150.1 & 2.8 & \textbf{3.8} & 3.1\\ \hline
	\end{tabular}
	\end{center}
	%\vspace{12pt}
\noindent Timing results (for GraphMat SSSP on dota-league)
	\begin{itemize}
		\item Finished file read of dota-league. time: 0.9495
		\item load graph: 3.48167 sec
		\item initialize engine: 3.60012e-05 sec
		\item run algorithm: 0.168526 sec
		\item print output: 0.105846 sec
		\item deinitialize engine: 6.00815e-05 sec
	\end{itemize}
\end{table}

@@

However, perusal of the log files reveals a different story. The listing shown below Table~\ref{tab:graphalytics-realworld} is the output from GraphMat itself. Graphalytics reports a 3.8 second runtime but 0.94 seconds of which GraphMat is simply reading the input file from disk. However, the GraphBIG timing (3.0 seconds) does not include the time to read the dota-league file. \textbf{If the time to read in the text file was computed correctly, GraphMat would be faster than GraphBIG.} To call this a fair comparison is dubious at best when certain phases of execution are omitted for some packages but not others.

@@

\section{Methodology}

@@

Our framework breaks the process of characterizing performance into five principal phases shown in \figurename~\ref{fig:epg-overview}, each of which requires no more than a single shell command. The five phases are as follows.
\begin{enumerate}
	\item Installing homogenized, stable forks of each software package to ensure homogeneity.
	\item Given a synthetic problem size or a real-world graph file, generate the files necessary to run each software package.
	\item Given a graph and the number of threads, run each algorithm using each software package multiple times.
	\item Parse through the log files to compress the output into a CSV.
	\item Analyze the data using the provided R scripts to generate plots.
\end{enumerate}

@@

\subsection{Datasets}
Some software packages provide serialized file formats to speed up file IO. Homogenizing the datasets creates extra copies of the graph files.

We refer to a graph's \emph{scale} when describing the size of the graph. Specifically, a graph with scale $S$ has $2^S$ vertices. For example, many of our experiments were performed on graphs with $2^{22} = 4,194,304$ vertices and an average of 16 edges per vertex. We measure strong scaling and speedup by varying the number of threads from one to the total number of threads available on our research server, 72.

@@

The Graphalytics results in Table~\ref{tab:graphalytics-realworld} were performed on the Dota-League dataset. This dataset contains 61,670 vertices and 50,870,313 edges and models interactions between players in the online video game Defense of the Ancients. This was sourced from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

@@

This study explores four shared memory parallel graph processing platforms and one distributed memory framework. The first three are so-called ``reference implementations'' while the remaining two are included because of their performance and popularity. Other popular libraries such as the Parallel Boost Graph Library~\cite{Gregor:2005:PBGL} are not considered here because the authors do not provide reference implementations. We consider the following systems.
\begin{enumerate}
	\item The Graph500~\cite{Murphy:2010:Graph500}, the canonical BFS benchmark which consists of a specification and reference implementation. We use a modified version most similar to 2.1.4. The Graph500 provides reference implementations for MPI and OpenMP but since our framework focuses on shared memory we use only the OpenMP version. The Graph500 uses a compressed sparse row (CSR) representation.
	\item The Graph Algorithm Platform (GAP) Benchmark Suite~\cite{Beamer:2015:GAPBench}, a set of reference implementations for shared memory graph processing. The author of GAP has contributed to the Graph500 so the BFS implementations are similar. Additionally, GAP uses OpenMP to achieve parallelism and uses a CSR representation.
	\item GraphBIG~\cite{Nai:2015:Graphbig} benchmark suite. We consider only the shared memory solutions but GraphBIG also provides GPU benchmarks. GraphBIG uses a CSR representation for graphs and OpenMP for parallelism.
	\item GraphMat~\cite{Sundaram:2015:GraphMat}, a library and programming model along with reference implementations of common algorithms. GraphMat uses a doubly-compressed sparse row representation and OpenMP for parallelism.
	\item PowerGraph~\cite{Gonzalez:2012:Powergraph}, a library and programming model for distributed (and shared memory) graph-parallel computation. PowerGraph includes additional toolkits for tasks such as clustering and computer vision. Parallelism is achieved via a combination of OpenMP and light-weight, user-level threads called fibers. Powergraph uses a novel storage scheme on top of CSR.
\end{enumerate}

Our approach is not specific or limited to these graph packages and can be extended to others. When comparing to Graphalytics, we use the most recent stable release, v0.2.

@@

We consider three algorithms: Breadth First Search (BFS), Single Source Shortest Paths (SSSP), and PageRank, though not all algorithms are implemented on all systems. We inspected the source code of the surveyed parallel graph processing systems to ensure the same phases of execution are measured across differing execution and programming paradigms.

@@

We select SSSP because of the straightforward extension from BFS; we need not modify the graph nor the root vertices from BFS. Furthermore, SSSP is used as a building block for other graph algorithms such as Betweenness Centrality.

@@

\subsection{Parsing and Data Analysis}
Our example analyzes data using the R programming language and shell scripting parsers. The example workflow provided in our framework yields the figures and tables generated in this paper.

@@

\subsection{Runtime for Synthetic Graphs}

\todo{redo this table}
\begin{table}
	\caption{Graphalytics on the same Kronecker graph with scale $22$ as used in other experiments. Performance results are in seconds with 32 threads. Community detection uses label propagation.}
	\centering
	\pgfplotstabletypeset[
	col sep=comma,
	columns={[index]0,graphmat,openg,powergraph},
	every head row/.style={after row=\midrule},
	columns/0/.style={string type, column type={l|}, column name={\large{\emph{Graphalytics}}}},
	columns/graphmat/.style={
		column name={GraphMat},
		string replace={0}{},
		dec sep align,
		empty cells with={N/A}
	},
	columns/openg/.style={
		column name={GraphBIG},
		dec sep align},
	columns/powergraph/.style={
		column name={PowerGraph},
		dec sep align}
	]{../report/runtime.csv}
	\label{tab:graphalytics}
\end{table}

In Table~\ref{tab:graphalytics} we show the results from running Graphalytics on a Kronecker graph of scale $22$ (4,194,304 vertices and approximately $16 \times 2^{22} \approx 33,500,000$ edges). An explanation of each algorithm is given in~\cite{Iosup:2016:Graphalyticstech}. Graphalytics by default does not perform SSSP on unweighted, undirected graphs. The discrepancy between PageRank values in Table~\ref{tab:graphalytics} and Fig.~\ref{fig:pr} is a result of the differing stopping criterion.

@@

\todo{What are the likely culprits for performance differences---measure cache miss rates. Try to find something which correlates with performance. Resource stalls will work easily (not as interesting) Cycles per instruction? Automate with perfexplorer or autoperf}
% Graphalytics also outputs MTEPS or millions of traversed edges per second. However, the graphalytics version does not make sense in all cases: for example, computing the local clustering coefficient involves traversing each edge multiple times (proportional to the sparsity of the graph), while BFS traverses each edge exactly once, and the number of edges traversed with PageRank depends on the connectivity of the graph and the number of iterations.

@@

\todo{Rotate the x-tick text so you can make all of the text larger}
\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_dsc.pdf}
	\end{minipage}
	\caption{The $y$-axes are logarithmic. The left box plot shows the time to compute BFS on 32 random roots while the right plot shows the times to construct the graph for each system. The Graph500 only constructs its graph once. GraphBIG reads in the file and generates the data structure simultaneously so is omitted.}
	\label{fig:bfs-time}
\end{figure}

@@

\subsection{Scalability}
Figures~\ref{fig:bfs-speedup} and~\ref{fig:bfs-efficiency} illustrate the parallel strong scaling of the BFS algorithm in different packages on a Kronecker graph of scale $20$ . The parallel speedup shown in Fig.~\ref{fig:bfs-speedup} is computed as $T_1 / T_n$ where $T_1$ is the sequential time and $T_n$ is the execution time on $n$ threads. 
\begin{figure}[htb]
	\centering
		% trim=left lower right upper
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 12pt, clip]{graphics/bfs_speedup20.pdf}
		\caption{Speedup of BFS for a scale-20 graph, black solid line represents ideal speedup. Both axes are logarithmic with an exception at 72 threads for readability. }
	\label{fig:bfs-speedup}
\end{figure}

@@

These plots show generally poor scaling for this size problem, a challenge facing most parallel graph algorithms. 
\todo[inline]{If time permits, replace with a larger graph to show scaling.}
In addition, $2^{20}$ vertices is a relatively ``small'' graph resulting for by today's standards and thus library designers focus on scalability for larger graphs. Another limitation may be the ability for OpenMP to efficiently handle such a large number of threads per machine. GCC version 4.8.5 supports OpenMP version 3.1 but the most recent release of OpenMP is 4.5.

\subsection{Real-World Datasets}
\begin{figure}
	\centering
		% trim=left lower right upper
		\includegraphics[width=\linewidth, trim=0 18pt 6pt 0pt, clip]{graphics/compare-realworld.pdf}
	\caption{Real world experiments using \emph{easy-parallel-graph-\textasteriskcentered}. The leftmost plot does is missing Powergraph because Powergraph does not provide BFS.}
	\label{fig:epg-realworld}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth, trim=0 18pt 6pt 0pt, clip]{graphics/graphalytics-GraphBIG.png}
	\caption{Real-world and synthetic experiments using Graphalytics on GraphBIG using 32 threads. Graphalytics outputs one HTML page per software package.}
	\label{fig:graphalytics-html}
\end{figure}
%\begin{table}
%	\centering
%	\caption{This table shows averages of runtimes for the Dota-League dataset using our method with 32 threads. We average over 32 runs and since the data structure construction times do not differ significantly we average across all algorithms. Both PowerGraph and GraphBIG read in the files and build a graph data structure simultaneously so we cannot measure data structure construction time.}
%	\begin{tabular}{l|c|c|c|c}
%		\multicolumn{2}{l}{\large{\emph{easy-parallel-graph\mbox{-\textasteriskcentered}}}} & & &\\
%		System & \begin{tabular}[x]{@{}c@{}}Construct Data \\ Structure\end{tabular}
%		& \begin{tabular}[x]{@{}c@{}}Run \\ BFS\end{tabular}
%		& \begin{tabular}[x]{@{}c@{}}Run \\ SSSP\end{tabular}
%		& \begin{tabular}[x]{@{}c@{}}Run \\ PageRank\end{tabular}
%		\\ \hline
%		GraphBIG   & N/A   & 0.01133        & {\bf 0.03791} & 19.5325  \\
%		PowerGraph & N/A   & N/A            & 0.3687        & 3.4406 \\
%		GAP        & 0.351 & 0.07449        & 0.0746        & {\bf 0.2465} \\
%		GraphMat   & 2.207 & {\bf 0.02273 } & 0.0397        & 1.6952 \\
%	\end{tabular}
%	\label{tab:epg-perf}
%\end{table}
%1         GAP      BFS Data structure build  0.25732000
%2    GraphMat      BFS Data structure build  2.10878359
%3         GAP PageRank Data structure build  0.32198688
%4    GraphMat PageRank Data structure build  2.15872441
%5         GAP     SSSP Data structure build  0.47479500
%6    GraphMat     SSSP Data structure build  2.35433178
%7         GAP      BFS                 Time  0.07449250
%8    GraphBIG      BFS                 Time  0.01132847
%9    GraphMat      BFS                 Time  0.02272806
%10        GAP PageRank                 Time  0.24646406
%11   GraphBIG PageRank                 Time 19.53251875
%12   GraphMat PageRank                 Time  1.69520531
%13 PowerGraph PageRank                 Time  3.44062500
%14        GAP     SSSP                 Time  0.07461594
%15   GraphBIG     SSSP                 Time  0.03791720
%16   GraphMat     SSSP                 Time  0.03936912
%17 PowerGraph     SSSP                 Time  0.36874997

>>>>

An overarching issue among these systems is the lack of comprehensive comparisons. One possible reason is the considerable effort involved in getting each system to run: satisfying dependencies and ensuring data are correctly formatted can be time consuming tasks. Moreover, some systems are developed with large, multi-node clusters in mind while others only work with shared memory, single-node computers. An example of the former is GraphX~\cite{Xin:2013:GraphX}, which incurs some overhead while gaining fault tolerance and an example of the latter is Ligra~\cite{Shun:2013:Ligra}, a framework requiring a shared-memory architecture. Additionally, there is a growing number of systems designed to run on GPUs \cite{Zhong:2014:Medusa}.
% Ligra's motivation for shared memory is the claim that `` the largest publicly available real-world graphs all fit in shared memory.''

@@

To address these issues, we introduce \emph{easy-parallel-\mbox{graph-\textasteriskcentered}},\footnote{Available at \url{https://github.com/HPCL/easy-parallel-graph}} a framework which simplifies the installation, comparison, and performance analysis of the three most widely implemented graph algorithm building blocks: breadth first search (BFS), single source shortest paths (SSSP), and PageRank (PR). With this framework we select a small number of graph processing libraries over which we homogenize all aspects of execution to compare performance fairly. Another contribution of this paper is a demonstration of the use of this framework to analyze the performance, energy consumption, and scalability of specific algorithm implementations. This enables algorithm designers to select high performance reference implementations with which they can compare their improvements. To be clear, we are not proposing a new benchmark suite or providing any new reference implementations. Instead, we introduce a method for detailed and fair comparison of existing graph software packages without requiring the user to be familiar with them.

Section~\ref{sec:relatedwork} describes previous performance analysis of parallel graph algorithms. Section~\ref{sec:meth} describes the architecture of \emph{easy-parallel-\mbox{graph-\textasteriskcentered}} and the algorithms, datasets, and graph processing implementations it uses. Section~\ref{sec:perf} presents performance analysis and discusses possible explanations for the results. Section~\ref{sec:fw} describes future directions for this research and is followed by the conclusion.

\section{Related Work}\label{sec:relatedwork}
Most performance analysis of graph processing systems come from the empirical results of each new library designer's publications. For example, GraphMat~\cite{Sundaram:2015:GraphMat}, PowerGraph~\cite{Gonzalez:2012:Powergraph}, and GraphBIG~\cite{Nai:2015:Graphbig} present a comparison of performance between their system and a selection of other software packages and in general any new approach will compare itself to existing implementations. We assume the performance results which accompany a new library are at risk of bias because the authors typically know their own package better than the competition and as such are aware of more optimizations. For a collection of references to the original papers of each new library or framework, see \cite{Doekemeijer:2015:GPFSurvey}. Instead, we focus on related research which provide analysis of existing software.

Nai et al.~\cite{Nai:2016:architectural} provide a detailed performance analysis using GraphBIG as their reference implementation. Their approach also compares GraphBIG to the GraphLab and Pregel execution models (Gather Apply Scatter). Their analysis considers architectural performance measurements such as cache miss rates to measure bottlenecks and computational efficiency for a variety of datasets. While incredibly thorough, these analyses focus on covering a wide range of datasets and appear difficult to apply to a new approach. For example, GraphMat reduces computation to sparse matrix operations which may not suffer from the same memory bottlenecks indicated in~\cite{Nai:2016:architectural}.

Research by Song et al.~\cite{Song:2016:LoadBalancing} and LeBeane et al.~\cite{LeBeane:2015:GraphHetero} considers the important problem of partitioning on heterogeneous architectures and provides profiling to reduce runtime, improve load balance, and reduce energy consumption.  Our framework instead focuses on shared memory packages.

Beyond the aforementioned performance results, a number of reports focus on performance analysis not tied to a particular software package. Satish et al.~\mbox{\cite{Satish:2014:NavigatingGraph}} analyze the performance of several systems on datasets on the order of $30$ billion edges and~\cite{Lu:2014:ExperimentalEval} uses six real-world datasets and focuses on the vertex-centric programming model. These implementations are hand tuned and provide recommendations for future improvements.

\begin{table}
	\caption{Graphalytics: tabulated sample run times (seconds) with 32 threads. Just one run per experiment is performed. Below the table is an excerpt from the GraphMat log file.}
	\label{tab:graphalytics-realworld}
	\vspace{-10pt}
	\begin{center}
		% cit-Patents: (3 774 768 vertices, 16 518 948 edges)
		% dota-league (61 170 vertices, 50 870 313 edges)
		\begin{tabular}{l|c|c|c|c|c|c}
			%% BN: Commented out because it's redundant with the caption
			%\multicolumn{5}{l}{\large{\emph{Graphalytics}}}  \\
			GraphBIG & BFS  & CDLP  &   LCC   &  PR &  SSSP & WCC \\ \hline
			cit-Patents & 0.8 & 11.8 & 15.5  &  4.5 & N/A  & 1.3 \\
			dota-league & 1.1 & 3.9 & 1073.7 &  {\bf 2.6} & 3.0 & 1.0 \\
			\multicolumn{7}{l}{} \\
			PowerGraph & BFS  & CDLP  &   LCC   &  PR &  SSSP & WCC \\ \hline
			cit-Patents & 13.8 & 30.1 & 23.9 & 18.8 & N/A & 22.1 \\
			dota-league & 25.6 & 31.2 & 458.1 & 26.7 & 28.9 & 22.9 \\
			% graph500-22 43.0 s 55.6 s 299.8 s 46.4 s N/A    40.5 s
			\multicolumn{7}{l}{} \\
			GraphMat & BFS  & CDLP  &   LCC   &  PR &  SSSP & WCC \\ \hline
			cit-Patents &  7.5 &  20.1 & 9.8 &  8.1 & N/A  & 6.6 \\
			dota-league &  2.7 &  21.2 & 239.7 & {\bf 6.3} & 9.4 & 6.9 \\
		\end{tabular}
	\end{center}

	\noindent Timing results (for GraphMat PageRank on dota-league)
	\begin{itemize}
		\item Finished file read of dota-league. time: 2.65211
		\item load graph: 5.91229 sec
		\item initialize engine: 8.32081e-05 sec
		\item run algorithm 1 (count degree): 0.0555639 sec
		\item run algorithm 2 (compute PageRank): 0.149445 sec
		\item print output: 0.0641179 sec
		\item deinitialize engine: 0.00022006 sec
	\end{itemize}
\end{table}

@@

@@

However, perusal of the log files reveals a different story. The bullets shown below Table~\ref{tab:graphalytics-realworld} summarize the output from GraphMat itself. Graphalytics reports a 6.3 second runtime but 2.7 seconds of that time GraphMat is simply reading the input file from disk. However, the GraphBIG timing (2.6 seconds) does not include the time to read the dota-league file. \textbf{If the time to read in the text file was ignored then GraphMat would complete nearly twice as quickly.} To call this a fair comparison is dubious at best when certain phases of execution are omitted for some packages but not others.

@@

\section{Methodology}\label{sec:meth}

@@

Our framework breaks the process of characterizing performance into five principal phases shown in \figurename~\ref{fig:epg-overview}, each of which requires no more than a single shell command. The five phases are as follows.
\begin{enumerate}
	\item Installing modified, stable forks of each software package to ensure homogeneity.
	\item Given a synthetic graph size or a real-world graph file, generate the files necessary to run each software package.
	\item Given a graph and the number of threads, run each algorithm using each software package multiple times.
	\item Parse through the log files to compress the output into a CSV.
	\item Analyze the data using the provided R scripts to generate plots.
\end{enumerate}

@@

\subsection{Datasets}
Homogenizing the datasets creates copies of the graph files and auxiliary files in various formats. This is both to ensure they are correctly formatted for each system and to speed up file I/O whenever possible by using the library designer's serialized data structure file formats.

We refer to a graph's \emph{scale} when describing the size of the graph. Specifically, a graph with scale $S$ has $2^S$ vertices. For example, many of our experiments were performed on graphs with $2^{22} = 4,194,304$ vertices and an average of 16 edges per vertex. We measure parallel efficiency and speedup by varying the number of threads from one to the total number of threads available on our research server, 72.

@@

Part of the Graphalytics results in Table~\ref{tab:graphalytics-realworld} were performed on the Dota-League dataset. We use this dataset for other experiments and as well. It contains 61,670 vertices and 50,870,313 edges and models interactions between players in the online video game Defense of the Ancients. This was sourced from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

@@

This study explores four shared memory parallel graph processing frameworks and one distributed memory framework operating on a single node (Powergraph). The first three are so-called ``reference implementations'' while the remaining two are included because of their performance and popularity. Other popular libraries such as the Parallel Boost Graph Library~\cite{Gregor:2005:PBGL} are not considered here because the authors do not provide reference implementations. The frameworks are as follows:
\begin{enumerate}
	\item The Graph500~\cite{Murphy:2010:Graph500}, the canonical BFS benchmark which consists of a specification and reference implementation. We use a modified version most similar to 2.1.4. The Graph500 provides reference implementations for MPI and OpenMP but since our framework focuses on shared memory we use only the OpenMP version. The Graph500 uses a compressed sparse row (CSR) representation.
	\item The Graph Algorithm Platform (GAP) Benchmark Suite~\cite{Beamer:2015:GAPBench}, a set of reference implementations for shared memory graph processing. The author of GAP has contributed to the Graph500 so the BFS implementations are similar. Additionally, GAP uses OpenMP to achieve parallelism and uses a CSR representation.
	\item GraphBIG~\cite{Nai:2015:Graphbig} benchmark suite. We consider only the shared memory solutions but GraphBIG also provides GPU benchmarks. GraphBIG uses a CSR representation for graphs and OpenMP for parallelism.
	\item GraphMat~\cite{Sundaram:2015:GraphMat}, a library and programming model along with reference implementations of common algorithms. GraphMat uses a doubly-compressed sparse row representation and OpenMP for parallelism.
	\item PowerGraph~\cite{Gonzalez:2012:Powergraph}, a library and programming model for distributed (and shared memory) graph-parallel computation. PowerGraph includes additional toolkits for tasks such as clustering and computer vision. Parallelism is achieved via a combination of OpenMP and light-weight, user-level threads called fibers. Powergraph uses a novel storage scheme on top of CSR.
\end{enumerate}

Our approach is not specific or limited to these graph packages and can be extended to others. When comparing to Graphalytics, we use the most recent stable release, v0.3.

@@

We consider three parallel algorithms: Breadth First Search (BFS), Single Source Shortest Paths (SSSP), and PageRank, though not all algorithms are implemented on all systems. We inspected the source code of the surveyed parallel graph processing systems to ensure the same phases of execution are measured across differing execution and programming paradigms.

@@

We select SSSP because of the straightforward extension from BFS; we need not modify the graph and can use the same the root vertices from BFS. Furthermore, SSSP is used as a building block for other graph algorithms such as Betweenness Centrality.

@@

\subsection{Parsing and Data Analysis}
Our example analyzes data using the R programming language and shell (Bash and AWK) parsers. The example workflow provided in our framework yields the figures and tables generated in this paper.

@@

\subsection{Runtime for Synthetic Graphs}
\begin{table}
	\caption{Graphalytics on the same Kronecker graph with scale $22$ as used in other experiments. Performance results are in seconds with 32 threads. Community detection uses label propagation.}
	\centering
	\pgfplotstabletypeset[
	col sep=comma,
	columns={[index]0,graphmat,openg,powergraph},
	every head row/.style={after row=\midrule},
	columns/0/.style={string type, column type={l|}, column name={\large{\emph{Graphalytics}}}},
	columns/graphmat/.style={
		column name={GraphMat},
		string replace={0}{},
		dec sep align,
		empty cells with={N/A}
	},
	columns/openg/.style={
		column name={GraphBIG},
		dec sep align},
	columns/powergraph/.style={
		column name={PowerGraph},
		dec sep align}
	]{../report/runtime.csv}
	\label{tab:graphalytics}
\end{table}

In Table~\ref{tab:graphalytics} we show the results from running Graphalytics on a Kronecker graph of scale $22$ (4,194,304 vertices and approximately $16 \times 2^{22} \approx 33,500,000$ edges). An explanation of each algorithm is given in~\cite{Iosup:2016:Graphalyticstech}. Graphalytics by default does not perform SSSP on unweighted, undirected graphs. The discrepancy between PageRank values in Table~\ref{tab:graphalytics} and Fig.~\ref{fig:pr} is a result of the differing stopping criterion and the aforementioned inconsistency of Graphalytics's performance collection scheme.

@@

% \todo{What are the likely culprits for performance differences---measure cache miss rates. Try to find something which correlates with performance. Resource stalls will work easily (not as interesting) Cycles per instruction? Automate with perfexplorer or autoperf}
% Graphalytics also outputs MTEPS or millions of traversed edges per second. However, the graphalytics version does not make sense in all cases: for example, computing the local clustering coefficient involves traversing each edge multiple times (proportional to the sparsity of the graph), while BFS traverses each edge exactly once, and the number of edges traversed with PageRank depends on the connectivity of the graph and the number of iterations.

@@

\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 36pt 18pt 0, clip]{graphics/bfs_dsc.pdf}
	\end{minipage}
	\caption{The $y$-axes are logarithmic. The left box plot shows the time to compute BFS on 32 random roots while the right plot shows the times to construct the graph for each system. The Graph500 only constructs its graph once. GraphBIG reads in the file and generates the data structure simultaneously so is omitted.}
	\label{fig:bfs-time}
\end{figure}

@@

\subsection{Scalability}
Figures~\ref{fig:bfs-speedup} and~\ref{fig:bfs-efficiency} illustrate the parallel strong efficiency of the BFS algorithm in different packages on a Kronecker graph of scale $20$ . The parallel speedup shown in Fig.~\ref{fig:bfs-speedup} is computed as $T_1 / T_n$ where $T_1$ is the sequential time and $T_n$ is the execution time on $n$ threads. 
\begin{figure}[htb]
	\centering
		% trim=left lower right upper
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 12pt, clip]{graphics/bfs_speedup20.pdf}
		\caption{Speedup of BFS for a scale-20 graph, black solid line represents ideal speedup. Both axes are logarithmic with an exception at 72 threads for readability. }
	\label{fig:bfs-speedup}
\end{figure}

@@

These plots show generally poor scaling for this size problem, a challenge facing most parallel graph algorithms. In addition, $2^{20}$ vertices is a relatively small graph by today's standards and thus library designers focus on scalability for larger graphs. Another limitation may be the ability for OpenMP to efficiently handle such a large number of threads per machine. GCC version 4.8.5 supports OpenMP version 3.1 but the most recent release of OpenMP is 4.5.

\subsection{Real-World Datasets}
\begin{figure}
	\includegraphics[width=\linewidth, trim=0 18pt 6pt 0pt, clip]{graphics/graphalytics-GraphBIG.png}
	\caption{Real-world and synthetic experiments using Graphalytics on GraphBIG using 32 threads. Graphalytics outputs one HTML page per software package.}
	\label{fig:graphalytics-html}
\end{figure}
%\begin{table}
%	\centering
%	\caption{This table shows averages of runtimes for the Dota-League dataset using our method with 32 threads. We average over 32 runs and since the data structure construction times do not differ significantly we average across all algorithms. Both PowerGraph and GraphBIG read in the files and build a graph data structure simultaneously so we cannot measure data structure construction time.}
%	\begin{tabular}{l|c|c|c|c}
%		\multicolumn{2}{l}{\large{\emph{easy-parallel-graph\mbox{-\textasteriskcentered}}}} & & &\\
%		System & \begin{tabular}[x]{@{}c@{}}Construct Data \\ Structure\end{tabular}
%		& \begin{tabular}[x]{@{}c@{}}Run \\ BFS\end{tabular}
%		& \begin{tabular}[x]{@{}c@{}}Run \\ SSSP\end{tabular}
%		& \begin{tabular}[x]{@{}c@{}}Run \\ PageRank\end{tabular}
%		\\ \hline
%		GraphBIG   & N/A   & 0.01133        & {\bf 0.03791} & 19.5325  \\
%		PowerGraph & N/A   & N/A            & 0.3687        & 3.4406 \\
%		GAP        & 0.351 & 0.07449        & 0.0746        & {\bf 0.2465} \\
%		GraphMat   & 2.207 & {\bf 0.02273 } & 0.0397        & 1.6952 \\
%	\end{tabular}
%	\label{tab:epg-perf}
%\end{table}
%1         GAP      BFS Data structure build  0.25732000
%2    GraphMat      BFS Data structure build  2.10878359
%3         GAP PageRank Data structure build  0.32198688
%4    GraphMat PageRank Data structure build  2.15872441
%5         GAP     SSSP Data structure build  0.47479500
%6    GraphMat     SSSP Data structure build  2.35433178
%7         GAP      BFS                 Time  0.07449250
%8    GraphBIG      BFS                 Time  0.01132847
%9    GraphMat      BFS                 Time  0.02272806
%10        GAP PageRank                 Time  0.24646406
%11   GraphBIG PageRank                 Time 19.53251875
%12   GraphMat PageRank                 Time  1.69520531
%13 PowerGraph PageRank                 Time  3.44062500
%14        GAP     SSSP                 Time  0.07461594
%15   GraphBIG     SSSP                 Time  0.03791720
%16   GraphMat     SSSP                 Time  0.03936912
%17 PowerGraph     SSSP                 Time  0.36874997

In addition to the data presented in Table~\ref{tab:graphalytics}, one may want to know the cost of performing additional computations on a graph once the data structures have been built. Figure~\ref{fig:epg-realworld} compiles average performance results for \emph{easy-parallel-graph-\textasteriskcentered}. We contrast this with Graphalytics which outputs an HTML page resulting from a single trial. We display a partial screenshot in \figurename~\ref{fig:graphalytics-html}.

\begin{figure}
	\centering
	% trim=left lower right upper
	\includegraphics[width=\linewidth, trim=0 18pt 6pt 0pt, clip]{graphics/compare-realworld.pdf}
	\caption{Real world experiments using \emph{easy-parallel-graph-\textasteriskcentered}. The leftmost plot does is missing Powergraph because Powergraph does not provide BFS.}
	\label{fig:epg-realworld}
\end{figure}
